### 2014-05-15 patchy-selection-paper.tex

%%%%%%%%%%%%
\subsection{Multiple patches}

We can extend this reasoning to multiple patches of unadapted habitat 
in which the $B$ allele can facilitate local adaptation. 
Under our assumptions of mutational exclusion and rapid establishment, 
each patch will adapt through establishment of a single copy of the $B$ allele, 
either by migration or mutation.
If we imagine that the times between successive establishments is the result of many nearly-independent attempts
with small probability of success,
the process of occupation is well approximated by a discrete-state Markov chain,
whose state is the labeling of patches according to the colonizing allele
(or none, for those that haven't yet adapted).

If not yet adapted,
a patch of area $A$ will acquire the adaptation by new mutation at rate $2 s_p \mu \rho A/\xi^2$.
Without loss of generality, assume that patches $1, \ldots, k$ are already adapted,
and these patches are at distances $r_1, \ldots, r_k$ away from this unadapted patch, 
then the total rate of adaptation through migration is approximately
\begin{equation}
  \frac{ s_p \rho }{s_m \xi^2} \sum_{i=1}^{k} \exp\left(- r_i \sqrt{2 s_m} /\sigma\right),
\end{equation}
and the probability that the migrant, adapted allele 
comes from patch $i$ is 
\begin{align}
  \exp\left(- r_i \sqrt{2 s_m} /\sigma\right)/\sum_{j=1}^{k} \exp\left(- r_j \sqrt{2 s_m} /\sigma\right) .
\end{align}
These specify the transition rates of the process.

@@

If a patch adapts through new mutation, the genomic region surrounding the selected site will hitchhike \citep{maynardsmith1974hitchhiking,kaplan1989hitchhiking} along with it,
so that at least initially, all adapted individuals share a fairly long stretch of haplotype.
This association gets slowly whittled down by recombination during interbreeding at the edge of the patch;
but there will always be longer LD nearby to the selected site.
\plr{Could work out on what time scale it gets whittled down and the shared length at stationarity.}

On the other hand, if an already adapted patch colonizes another through migration,
then the newly colonized patch will inherit a long piece of haplotype around the selected site from the originating patch.
The length of this haplotype is roughly inversely proportional to the number of generations the allele spends ``in transit'',
because each reproductive event in locations where the allele is at low frequency
recombines haplotypes of the $b$ allele closer to the adaptive allele.
(In this case, the haplotype is literally hitchhiking across geography!)

As noted above, the length of the hitchhiking segment depends on the time
spent ``in transit'' between the patches.
Above we argued that a good model for this transit is a single Brownian trunk lineage
surrounded by a cloud of close relatives of random size $K$
whose chance of surviving until $t$ generations is $p_e(t) \simeq e^{-\alpha t} / \E[K]$.
Consider a single such family, and let $\tau$ be the (random) time it first hits the new patch,
with $\tau = \infty$ if it never does.
By pulling out the constant $1/\E[K]$, we can use standard results on hitting times of $d$-dimensional Brownian motion
that is killed at rate $\alpha$.
In particular, we know that if the patch is circular with radius $w$, then
using \citet[][(2.2.0.1 and 4.2.0.1)]{borodin2002handbook}
\begin{align}
  \E[e^{-\ell \tau}] &=
    \begin{cases}
      e^{- R \sqrt{2(\alpha+\ell)}/\sigma} \qquad & d=1 \\
      \frac{ K_0( (R+w)\sqrt{2(\alpha+\ell)}/\sigma) }{ K_0( w\sqrt{2(\alpha+\ell)}/\sigma) } \qquad & d=2 
    \end{cases} 
    % &= \left( R/w +1 \right)^{-\frac{d-1}{2}} e^{- R \sqrt{2(\alpha+\ell)}/\sigma}  (1+O(1/w)) .
\end{align}
In particular, we can find 
$\P\{\tau < \infty\} = \lim_{\ell \to 0} \E[\exp(-\ell \tau)]$. % \simeq (R/w+1)^{-(d-1)/2} \exp(- R \sqrt{2\alpha}/\sigma)$.

@@

Since each generation spent in transit provides an opportunity for recombination,
if recombination is Poisson, the length of the hitchhiking segment (in Morgans) on each side of the site will be exponentially distributed
with mean $\tau$, so that if $L$ is the length on, say, the right, then in one dimension
\begin{align} \label{eqn:haplen_cdf}
\P\{L>\ell\} &= \E[e^{-\ell \tau}|\tau<\infty] \\
  &= \exp\left\{{-\frac{R}{\sigma}\left(\sqrt{2(\ell+\alpha)} - \sqrt{2\alpha}\right)}\right\} .
\end{align}
This is also the Laplace transform of $\tau$;
we can differentiate it to find that $\E[\tau] = (R\sqrt{2\alpha}/\sigma)\times 1/(2\alpha)$
and $\var[\tau] = (R\sqrt{2\alpha}/\sigma) \times 1/(2\alpha)^2$.
Furthermore, this implies that if $Y$ is an exponential random variable with rate $R\sqrt{2}/\sigma$,
then $L$ has the same distribution as $(Y + \sqrt{\alpha})^2 - \alpha$,
from which we can obtain that
$\E[L] = \sigma \sqrt{2\alpha}/R + \sigma^2/R^2$ and $\var[L] = 2\alpha\sigma^2/R^2 + 4 \sigma^3 \sqrt{2\alpha}/R^3 + 5 \sigma^4 / R^4$.
For two dimensions, asymptotics of Bessel functions show that $L$ has the same tail behavior,
but how other properties might change is unclear.

>>>>

%%%%%%%%%%%%
\subsection{Multiple patches}
We can extend this reasoning to multiple patches of unadapted habitat 
in which the $B$ allele can facilitate local adaptation. 
Under our assumptions of mutational exclusion and rapid establishment, 
each patch will adapt through establishment of a single copy of the $B$ allele, 
either by migration or mutation.
If we imagine that the times between successive establishments is the result of many nearly-independent attempts
with small probability of success,
the process of occupation is well approximated by a discrete-state Markov chain,
whose state is the labeling of patches according to the colonizing allele
(or none, for those that haven't yet adapted).

If not yet adapted, a patch of area $A$ will acquire the adaptation by new mutation at rate $2 s_p \mu \rho A/\xi^2$.
Without loss of generality, assume that patches $1, \ldots, k$ are already adapted,
and these patches are at distances $r_1, \ldots, r_k$ away from this unadapted patch, 
then the total rate of adaptation through migration is approximately
\begin{equation}
  \frac{ s_p \rho }{\alpha \xi^2} \sum_{i=1}^{k} \left(r_i \sqrt{2 s_m} /\sigma \right) \exp\left(- r_i \sqrt{2 s_m} /\sigma\right),
\end{equation}
and conditional on coming from migration the probability the adapted allele 
comes from patch $i$ is 
\begin{align}
  \frac{\left(r_i \sqrt{2 s_m} /\sigma \right) \exp\left(- r_i \sqrt{2 s_m}
    /\sigma\right)} {\sum_{j=1}^{k}  \left(r_i \sqrt{2 s_m} /\sigma \right) \exp\left(- r_i \sqrt{2 s_m}
    /\sigma\right) } .
\end{align}
These specify the transition rates of the process.

@@

If a patch adapts through new mutation or a rare migrant lineage, the
genomic region surrounding the selected site will hitchhike along with
it \citep{maynardsmith1974hitchhiking},
so that at least initially, all adapted individuals within a patch
share a fairly long stretch of haplotype. \gc{Pairs of individuals
  within a patch will share a haplotype of mean genetic length of
  $\approx \approx \log(\rho A s_p)/s_p$ around the selected site,
as long as the patch is reasonably well mixed by dispersal
\citep[otherwise see][]{BartonEtheridge13spatialHH}. }
This association gets slowly whittled down by recombination during interbreeding at the edge of the patch;
but there will always be longer LD nearby to the selected site. 
\plr{Could work out on what time scale it gets whittled down and the shared length at stationarity.}

When an already adapted patch colonizes another through migration,
then the newly colonized patch will inherit a long piece of haplotype around the selected site from the originating patch.
The genetic length of this haplotype is roughly inversely proportional to the
number of generations the allele spends ``in transit'', 
because each reproductive event in locations where the allele is at low frequency recombines haplotypes of the $b$ allele closer to the adaptive allele.
(In this case, the haplotype is literally hitchhiking across geography!)

Above we argued that a good model for this transit is a single Brownian trunk lineage
surrounded by a cloud of close relatives of random size $K$
whose chance of surviving until $t$ generations is $p_e(t) \simeq e^{-\alpha t} / \E[K]$.
Consider a single such family, and let $\tau$ be the (random) time it first hits the new patch,
with $\tau = \infty$ if it never does. The length of the hitchhiking segment depends on the time
spent by this trunk lineage ``in transit'' between the patches, i.e. $\tau$.

By pulling out the constant $1/\E[K]$, we can use standard results on hitting times of $d$-dimensional Brownian motion
that is killed at rate $\alpha$ \citet[][(2.2.0.1 and 4.2.0.1)]{borodin2002handbook} 
In particular, if the patch is circular with radius $w$ and a distance
$R$ from the already adapted patch, then we have the useful result
about the Laplace transform of $\tau$ our transit time
\begin{align}
  \E[e^{-\ell \tau}] &=
    \begin{cases}
      e^{- R \sqrt{2(\alpha+\ell)}/\sigma} \qquad & d=1 \\
      \frac{ K_0( (R+w)\sqrt{2(\alpha+\ell)}/\sigma) }{ K_0( w\sqrt{2(\alpha+\ell)}/\sigma) } \qquad & d=2 
    \end{cases} \label{borodinresult}
    % &= \left( R/w +1 \right)^{-\frac{d-1}{2}} e^{- R \sqrt{2(\alpha+\ell)}/\sigma}  (1+O(1/w)) .
\end{align}
\gc{define $K_0$, bessel? We've used K for family size}
We can find that the probability that a particular
trunk lineageis killed before it reaches the new patch is
$\P\{\tau < \infty\} = \lim_{\ell \to 0} \E \left[\exp(-\ell \tau)
\right]$. We are interested in lineages that
manage to establish on the patch i.e. $\tau< \infty$, and so in one dimension
\begin{equation} \label{eqn:haplen_cdf}
\E[e^{-\ell \tau}|\tau<\infty]  = \exp\left\{{-\frac{R}{\sigma}\left(\sqrt{2(\ell+\alpha)} - \sqrt{2\alpha}\right)}\right\} .
\end{equation}
We can differentiate this to find that our expected transit time is
\begin{equation}
\E[\tau] = (R\sqrt{2\alpha}/\sigma)\times 1/(2\alpha)
\end{equation}
and $\var[\tau] = (R\sqrt{2\alpha}/\sigma) \times 1/(2\alpha)^2$ in
one dimension.

% \simeq (R/w+1)^{-(d-1)/2} \exp(- R \sqrt{2\alpha}/\sigma)$.

@@

Since each generation spent in transit provides an opportunity for recombination,
if recombination is Poisson, the length of the haplotype (in Morgans)
initially shared between populations on each side of the site will be exponentially distributed
with mean $\tau$, so that if $L$ is the length of hitchhiking segment on, say, the right of
the selected locus, then
\begin{equation} \label{eqn:haplen_cdf2}
\P\{L>\ell\} = \E[e^{-\ell \tau}|\tau<\infty] 
%  &= \exp\left\{{-\frac{R}{\sigma}\left(\sqrt{2(\ell+\alpha)} - \sqrt{2\alpha}\right)}\right\} .
\end{equation}
which is our Laplace transform from above that in one dimension
gives us eqn. \eqref{eqn:haplen_cdf}. This implies that in one dimension, if $Y$ is an exponential random variable with rate $R\sqrt{2}/\sigma$,
then $L$ has the same distribution as $(Y + \sqrt{\alpha})^2 - \alpha$,
from which we can obtain that the expected length of shared hitchiking
haplotype is
\begin{equation} \label{eqn:haplotype_length}
\E[L] = \sigma \sqrt{2\alpha}/R + \sigma^2/R^2
\end{equation}
and thar $\var[L] = 2\alpha\sigma^2/R^2 + 4 \sigma^3 \sqrt{2\alpha}/R^3 + 5 \sigma^4 / R^4$.
For two dimensions, asymptotics of Bessel functions show that $L$ has the same tail behavior,
but how other properties might change is unclear. \gc{add some
  sentences of intuition of how parameters play into haplotype length
  here. is it really $\sigma \sqrt{\alpha}$ in \eqref{eqn:haplotype_length} and not $\sqrt{\alpha}/sigma$???}

\gc{Perhaps we should separate the above bit into things that are
  useful about $\tau$ and put those up by  \eqref{borodinresult} and
  things that are useful about $L$ here? We should also turn some of
  this into numbered equations. I tried to do this above, and have the
commented out original below}

% Since each generation spent in transit provides an opportunity for recombination,
% if recombination is Poisson, the length of the hitchhiking segment (in Morgans) on each side of the site will be exponentially distributed
% with mean $\tau$, so that if $L$ is the length of hitchhiking segment on, say, the right of
% the selected locus, then in one dimension
% \begin{align} \label{eqn:haplen_cdf}
% \P\{L>\ell\} &= \E[e^{-\ell \tau}|\tau<\infty] \\
%   &= \exp\left\{{-\frac{R}{\sigma}\left(\sqrt{2(\ell+\alpha)} - \sqrt{2\alpha}\right)}\right\} .
% \end{align}
% found by using eqn. \eqref{borodinresult}. This is also the Laplace transform of $\tau$;
% we can differentiate it to find that $\E[\tau] = (R\sqrt{2\alpha}/\sigma)\times 1/(2\alpha)$
% and $\var[\tau] = (R\sqrt{2\alpha}/\sigma) \times 1/(2\alpha)^2$.
% Furthermore, this implies that if $Y$ is an exponential random variable with rate $R\sqrt{2}/\sigma$,
% then $L$ has the same distribution as $(Y + \sqrt{\alpha})^2 - \alpha$,
% from which we can obtain that
% $\E[L] = \sigma \sqrt{2\alpha}/R + \sigma^2/R^2$ and $\var[L] = 2\alpha\sigma^2/R^2 + 4 \sigma^3 \sqrt{2\alpha}/R^3 + 5 \sigma^4 / R^4$.
% For two dimensions, asymptotics of Bessel functions show that $L$ has the same tail behavior,
% but how other properties might change is unclear.
